MAESTROeX: A Massively Parallel Low Mach Number Astrophysical Solver

### 2019-02-11 MAESTROeX_I/paper.tex
minor

\section{Governing Equations}
The derivation of the low Mach number hydrodynamics equations is given in Papers I-III.
The resulting equations are
\begin{eqnarray}
\frac{\partial(\rho X_k)}{\partial t} &=& -\nabla\cdot(\rho X_k\Ub) + \rho\omegadot_k,\label{eq:species}\\
\frac{\partial\Ub}{\partial t} &=& -\Ub\cdot\nabla\Ub  - \frac{1}{\rho}\nabla\pi - \frac{\rho-\rho_0}{\rho} g\eb_r,\label{eq:momentum}\\
\frac{\partial(\rho h)}{\partial t} &=& -\nabla\cdot(\rho h\Ub) + \frac{Dp_0}{Dt} + \rho\Hnuc + \rho\Hext,\label{eq:enthalpy}
\end{eqnarray}
where $\rho$, $\Ub$, and $h$ are the mass density,
velocity and specific enthalpy, respectively, and
$X_k$ are the mass fractions of species $k$ with associated
production rate $\omegadot_k$.  The species are constrained
such that $\sum_k X_k = 1$ giving $\rho = \sum_k (\rho X_k)$ and
\begin{equation}
\frac{\partial\rho}{\partial t} = -\nabla\cdot(\rho\Ub).
\end{equation}
The source terms $\Hext$ and $\Hnuc$ are the external heating rate and nuclear energy 
generation rate per unit mass.  The pressure is decomposed into a hydrostatic base state
 pressure, $p_0 = p_0(r,t)$, and a dynamic pressure, $\pi = \pi(\xb,t)$, such that 
$|\pi|/p_0 = \mathcal{O}(M^2)$ (we use $\xb$ to represent the Cartesian coordinate 
directions of the full state and $r$ to represent the radial coordinate direction for 
the base state).  We also define a base state density, $\rho_0 = \rho_0(r,t)$, 
which is in hydrostatic equilibrium with $p_0$, i.e., 
$\nabla p_0 = -\rho_0 g\eb_r$, where $g=g(r,t)$ is
the magnitude of the gravitational acceleration and $\eb_r$ is the unit vector in the
outward radial direction.

Mathematically, this system must still be closed by the equation of state which we
express as a divergence constraint on the velocity field (see Paper III),
\begin{equation}
\nabla\cdot(\beta_0\Ub) = \beta_0\left(S - \frac{1}{\gammaonebar p_0}\frac{\partial p_0}{\partial t}\right),\label{eq:U divergence}
\end{equation}
\MarginPar{This is different from the Jacobs 2016 paper where we switched to alt\_energy\_fix.}
where $\beta_0$ is a density-like variable that carries background stratification, defined as
\begin{equation}
\beta_0(r,t) = \rho_0(0,t)\exp\left(\int_0^r\frac{1}{\gammaonebar p_0}\frac{\partial p_0}{\partial r'}dr'\right),
\end{equation}
where $\gammaonebar$ is the lateral average of $\Gamma_1 = d(\log p)/d(\log\rho)$ at constant entropy.
The expansion term, $S$, incorporates local compressibility effects due to heat release from reactions, compositional changes, and external sources,
\begin{equation}
S = -\sigma\sum_k\xi_k\omegadot_k + \frac{1}{\rho p_\rho}\sum_k p_{X_k}\omegadot_k + \sigma\Hnuc + \sigma\Hext,\label{eq:S}
\end{equation}
where $p_{X_k} \equiv \left. \partial p / \partial X_k
\right|_{\rho,T,X_{j,j\ne k}}$, $\xi_k \equiv \left. \partial h /
\partial X_k \right |_{p,T,X_{j,j\ne k}},
p_\rho \equiv \left.
\partial p/\partial \rho \right |_{T, X_k}$, and $\sigma \equiv
p_T/(\rho c_p p_\rho)$, with $p_T \equiv \left. \partial p / \partial
T \right|_{\rho, X_k}$ and $c_p \equiv \left.  \partial h / \partial T
\right|_{p,X_k}$ is the specific heat at constant pressure.

@@

This work was supported by SciDAC... (contract number)

This research used resources of the National Energy Research Scientific Computing Center (NERSC), a U.S. Department of Energy Office of Science User Facility operated under Contract No. DE-AC02-05CH11231.

>>>>

\section{Governing Equations}
The derivation of the low Mach number hydrodynamics equations is given in Papers I-III.
We take the standard equations of reacting, compressible flow, and recast the equation
of state as a divergence constraint on the velocity field.  This is done by taking
the Langrangian derivative of the equation of state for pressure as a function of the
density, composition, and enthalpy, enforcing the constraint that the pressure is
a prescribed function of altitutde and time based on the hydrostatic equilibrium condition,
and substituting the equations for species and enthalpy in the resulting Lagrangian derivative.
The resulting equations are
\begin{eqnarray}
\frac{\partial\Ub}{\partial t} &=& -\Ub\cdot\nabla\Ub  - \frac{1}{\rho}\nabla\pi - \frac{\rho-\rho_0}{\rho} g\eb_r,\label{eq:momentum}\\
\frac{\partial(\rho X_k)}{\partial t} &=& -\nabla\cdot(\rho X_k\Ub) + \rho\omegadot_k,\label{eq:species}\\
\frac{\partial(\rho h)}{\partial t} &=& -\nabla\cdot(\rho h\Ub) + \frac{Dp_0}{Dt} + \rho\Hnuc + \rho\Hext,\label{eq:enthalpy}
\end{eqnarray}
\begin{equation}
\nabla\cdot(\beta_0\Ub) = \beta_0\left(S - \frac{1}{\gammaonebar p_0}\frac{\partial p_0}{\partial t}\right).\label{eq:U divergence}
\end{equation}
Here $\rho$, $\Ub$, and $h$ are the mass density,
velocity and specific enthalpy, respectively, and
$X_k$ are the mass fractions of species $k$ with associated
production rate $\omegadot_k$.  The species are constrained
such that $\sum_k X_k = 1$ giving $\rho = \sum_k (\rho X_k)$ and
\begin{equation}
\frac{\partial\rho}{\partial t} = -\nabla\cdot(\rho\Ub).
\end{equation}
The source terms $\Hext$ and $\Hnuc$ are the external heating rate and nuclear energy 
generation rate per unit mass.  The pressure is decomposed into a hydrostatic base state
 pressure, $p_0 = p_0(r,t)$, and a dynamic pressure, $\pi = \pi(\xb,t)$, such that 
$|\pi|/p_0 = \mathcal{O}(M^2)$ (we use $\xb$ to represent the Cartesian coordinate 
directions of the full state and $r$ to represent the radial coordinate direction for 
the base state).  We also define a base state density, $\rho_0 = \rho_0(r,t)$, 
which is in hydrostatic equilibrium with $p_0$, i.e., 
$\nabla p_0 = -\rho_0 g\eb_r$, where $g=g(r,t)$ is
the magnitude of the gravitational acceleration and $\eb_r$ is the unit vector in the
outward radial direction.

Mathematically, equations (\ref{eq:momentum})-(\ref{eq:enthalpy}) must still be closed by the equation of state which we
express as a divergence constraint on the velocity field (\ref{eq:U divergence}),
\MarginPar{This is different from the Jacobs 2016 paper where we switched to alt\_energy\_fix.}
where $\beta_0$ is a density-like variable that carries background stratification, defined as
\begin{equation}
\beta_0(r,t) = \rho_0(0,t)\exp\left(\int_0^r\frac{1}{\gammaonebar p_0}\frac{\partial p_0}{\partial r'}dr'\right),
\end{equation}
where $\gammaonebar$ is the lateral average of $\Gamma_1 = d(\log p)/d(\log\rho)$ at constant entropy.
The expansion term, $S$, incorporates local compressibility effects due to heat release from reactions, compositional changes, and external sources,
\begin{equation}
S = -\sigma\sum_k\xi_k\omegadot_k + \frac{1}{\rho p_\rho}\sum_k p_{X_k}\omegadot_k + \sigma\Hnuc + \sigma\Hext,\label{eq:S}
\end{equation}
where $p_{X_k} \equiv \left. \partial p / \partial X_k
\right|_{\rho,T,X_{j,j\ne k}}$, $\xi_k \equiv \left. \partial h /
\partial X_k \right |_{p,T,X_{j,j\ne k}},
p_\rho \equiv \left.
\partial p/\partial \rho \right |_{T, X_k}$, and $\sigma \equiv
p_T/(\rho c_p p_\rho)$, with $p_T \equiv \left. \partial p / \partial
T \right|_{\rho, X_k}$ and $c_p \equiv \left.  \partial h / \partial T
\right|_{p,X_k}$ is the specific heat at constant pressure.

@@

The work at LBNL was supported by the U.S. Department of Energy's Scientific Discovery Through Advanced Computing (SciDAC) program under contract No. DE-AC02-05CH11231.
This research used resources of the National Energy Research Scientific Computing Center (NERSC), a U.S. Department of Energy Office of Science User Facility operated under Contract No. DE-AC02-05CH11231.
